Selection on early survival does not explain germination rate clines in *Mimulus cardinalis*

### 2022-07-10 ms/ms.Rmd
verified make
archive

---
title: 'Selection on early survival does not explain germination rate clines in *Mimulus cardinalis* (Phrymaceae)'
author: "Christopher D. Muir$^{1,2,*}$, Courtney L. Van Den Elzen$^{1,3}$, Amy L. Angert$^{1,4}$"
output:
  bookdown::pdf_document2:
    includes:
      before_body: ../template/doc_prefix.tex
      in_header: ../template/preamble.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc: no
  bookdown::html_document2:
    number_sections: no
    theme: readable
    toc: yes
  bookdown::tufte_html2:
    number_sections: no
    toc: yes
  bookdown::word_document2: null
fontsize: 12pt
linestretch: 1.5
link-citations: yes
csl: ../template/american-journal-of-botany.csl
bibliography: ../ms/mimulus-germination.bib
---

@@

<!--- broad problem: local adaptation --->
Populations within species differ in ecologically important traits that often result from adaptation to different local environments [@turesson_genotypic_1922; @clausen_experimental_1948]. However, it is still rare that we understand the traits and selective agents underlying local adaptation [@wadgymar_identifying_2017]. The classic signature of local adaptation is crossing reaction norms for fitness measured in a reciprocal transplant experiment [@kawecki_conceptual_2004; @johnson_reciprocal_2021], such that populations have  higher relative fitness in their local environment but lower relative fitness in a foreign environment. For example, annual inland ecotypes of the yellow Monkeyflower *Mimulus guttatus* have higher relative fitness in seasonally dry sites compared to perennial coastal ecotypes that are adapted to year-round water availability. The relative fitness of coastal and inland ecotypes is reversed on the coast because of life history and other genetic differences between ecotypes [@lowry_widespread_2010]. The prevalence of local adaptation implies that selection varies spatially, populations have heritable variation in fitness, and selection is stronger than migration or drift. Exceptional cases of local adaptation over short temporal or spatial scales [e.g. @wright_indirect_2013; @grant_40_2014; @richardson_microgeographic_2014; @lescak_evolution_2015; @barrett_linking_2019; @divittorio_natural_2020] likely contribute to an availability bias that leads many to perceive that local adaptation is ubiquitous. However, systematic meta-analyses reveal that local adaptation is often weak or nonexistent [@leimu_meta-analysis_2008; @hereford_quantitative_2009; @brady_understanding_2019] and may be getting weaker because of anthropogenic climate change [@bontrager_climate_2020]. Populations may not be locally adapted if differential selection is weak relative to migration or drift or they lack heritable variation in traits under selection. Alternatively, recent anthropogenic climate change may have erased the signature of local adaptation by causing some foreign populations to have higher fitness than local populations [@bontrager_climate_2020]. For example, in the alpine plant *Boechera strica*, simulated climate change (mimicking early snow melt) favors populations from lower elevations and snow addition restores the relative fitness advantage of high-elevation populations [@anderson_climate_2020]. Understanding what causes trait variation among populations within species, whether adaptive or not, will help biologists better predict how populations will respond to environmental change.

@@

There are several experimental methods for measuring the agents and strength of selection to different environments under natural conditions [@wadgymar_identifying_2017], but here we focus on differential selection. One key prediction is that the direction of selection and position of the phenotypic optimum will differ between environments. If genetic variation in a trait *causes* fitness to vary, then genotypic selection analyses [@rausher_measurement_1992] should find that genotypes with high trait expression are favored in one environment but disfavored in another. For example, selection favors higher seed dormancy in southern environments of *Arabidopsis thaliana* and lower dormancy in northern environments, presumably to optimally time seedling establishment with the onset of favorable summer conditions [@postma_early_2016]. Populations can also simultaneously differ as a result of both adaptive and non-adaptive processes. When locally adapted populations differ in traits that do not themselves confer local adaptation, we expect population-level covariation between divergent traits and fitness but do not expect within-population variation to correlate with fitness. Hence, comparing the direction of among-genotype and among-population selection can help distinguish between phenotypic clines that are important for local adaptation and those that may have evolved nonadaptively.

@@

*Mimulus cardinalis* [$=$*Erythranthe cardinalis* @lowry_case_2019] is a perennial forb native to the Western US (California and Oregon). It is self-compatible, predominantly outcrossing, and hummingbird pollinated. We used five focal populations from throughout the geographic range of *M. cardinalis* (Table \ref{tab:focal_pops}; Fig. \ref{fig:range-map}). These five focal populations are a subset of those we used previously to identify a cline in germination and other traits [@muir_grow_2017]. Seeds were collected in the field from mature, undehisced fruits left in open coin envelopes for $2-4$ weeks to dry, then stored at room temperature. To minimize maternal effects, we grew a large number of field-collected seeds in the greenhouse and generated seed families for this experiment by hand-pollinating individuals using the breeding design described in the next section.

```{r, focal_pops, echo=FALSE, eval=TRUE}

tribble(
  ~`Source population`, ~Latitude, ~Longitude, ~`Elevation (mas)`,
  "Sweetwater River", "$32.900$", "$-116.585$", "$1180$",
  "West Fork Mojave River", "$34.284$", "$-117.378$", "$1120$",
  "North Fork Middle Tule River", "$36.201$", "$-118.759$", "$926$",
  "Little Jamison Creek", "$39.743$", "$-120.704$", "$1603$",
  "Rock Creek", "$43.374$", "$-122.957$", "$326$"
) |>
  knitr::kable(format = "markdown", caption = "\\label{tab:focal_pops} Source populations, including the name of the drainage where the seeds were collected, the latitude, longitude, and elevation in meters above sea level (mas).", align = "lccc", label = NA)
```

@@

$$V_G = 4~\sigma_{\mathrm{parent}}$$
Using $\sigma_{\mathrm{parent}}$ rather than $\sigma_{\mathrm{sire}}$ is more powerful because it uses all of the data. We estimated $\sigma_{\mathrm{parent}}$ and other parameters using a Bayesian mixed effects model. We describe the general approach here and provide detail specific to germination rate and survival below. We fit the model in *Stan* version $`r stringr::str_c(fit[["metadata"]]()["stan_version_major"], fit[["metadata"]]()["stan_version_minor"], fit[["metadata"]]()["stan_version_patch"], sep = ".")`$ [@stan_development_team_stan_2022] using **cmdstanr** version $`r packageVersion("cmdstanr")`$ [@gabry_cmdstanr_2022]. *Stan* calculates the posterior distribution of the model using the Hamiltonian Monte Carlo algorithm, which is similar to the more widely used Markov Chain Monte Carlo, but is faster and more efficient at sampling for many applications [@monnahan_faster_2017]. We used weakly informative normally distributed priors for parameters that affect the trait mean (intercepts and coefficients) and half-Cauchy priors for variance parameters. Weakly informative priors are strongly recommended for complex mixed models [@mcelreath_statistical_2016]. We ran $`r length(fit$metadata()$id)`$ parallel chains for $`r fit[["metadata"]]()[["iter_warmup"]]`$ warmup iterations, $`r fit[["metadata"]]()[["iter_sampling"]]`$ sampling iterations, and a thinning interval of $`r fit[["metadata"]]()[["thin"]]`$. This configuration allowed parameters to converge, which we defined as the convergence diagnostic $\hat{R} < 1.05$ [@vehtari_rank-normalization_2021]. We inspected posterior predictive plots to assess model adequacy using the `pp_check` function from the *R* package **bayesplot** version $`r packageVersion("bayesplot")`$ [@gabry_visualization_2019]. We used the posterior median for point estimates and calculated uncertainty with the $95$\% quantile intervals from the posterior distribution. *Stan* and *R* code are available on GitHub ([https://github.com/cdmuir/mimulus-germination](https://github.com/cdmuir/mimulus-germination)) and will be archived on Zenodo upon publication. Data will be deposited on Dryad upon publication.

@@

We estimated germination rate under benign greenhouse conditions by measuring time to germination, the number of days to visible emergence from the soil. There were $48$ randomized blocks split evenly between two time cohorts (North cohort: August $3 \textendash 4$ and South cohort: August $22 \textendash 24$, $2015$) in the University of British Columbia greenhouse (Vancouver, BC, Canada). The North seed cohort went into the North transplant garden (gardens described below) and the South seed cohort went into the South transplant garden. Seed cohorts were staggered so that seedlings were transplanted at the same ontogenetic stage, as described below. We distinguish between cohorts and gardens because germination for each transplant garden was assayed in the greenhouse prior to transplanting. Emergence is a reliable proxy for germination since seeds were sown directly on top of the soil, resulting in little delay between germination and visible cotyledons. We sowed $3 \textendash 5$ seeds per plug on moist perlite-peat moss potting medium (Sunshine #1, Sungro Horticulture, Agawam, Massachusetts, USA) and recorded the first day on which a plug had at least one germinated seedling. We later thinned each plug to one seedling by selecting the central-most individual, so the first germinated seedling recorded is generally not the same individual transplanted into the field. Hence, individual correlations between germination and survival are inappropriate and we analyze relationships between breeding values or population averages. We censused germination daily for most of the experiment, but census dates were more spread out later in the experiment when only a few plants remained. We accounted for census interval in our statistical model (see below). Gentle misting kept soil moist for the duration of the experiment. Most plugs ($`r length(which(data[["Germinate"]] == "Y"))`$ of $`r nrow(data)`$, $`r round(100 * length(which(data[["Germinate"]] == "Y")) / nrow(data), 1)`$%) had at least one germinant by the end of the experiment (North cohort: August 27, 2015; South cohort: September 20, 2015); we treated plants that did not germinate as missing data. $`r nRemove`$ ($`r round(100 * nRemove / length(which(data[["Germinate"]] == "Y")), 2)`$%) plants emerged before day $5$. We removed data on these individuals because they were likely not *M. cardinalis*, but rather contamination from another species' seed in the potting medium. Species can be difficult to tell apart as the cotyledons first emerge, but in plugs where a single germinant emerged before day $5$ it was always a contaminant species. It therefore seems most likely that all germinants before day $5$ were contamination, but observations during thinning showed that contamination was rare and it unlikely to have any substantial impact on our estimate of germination timing.

@@

After germination, we transplanted $20$ out of $24$ blocks of seedlings each to two experimental gardens located at the northern and southern portions of the species' range. We refer to these as the "North" and "South" gardens, respectively. We used some plants from the remaining $4$ blocks to replace those that did not germinate or died during transport. We transplanted seedlings at similar ontogenetic stages but temporally staggered. We planted the North and South cohorts $5 \textendash 6$ weeks after germination on September $9 \textendash 17$, $2015$ (North seed cohort into North garden) and October $2 \textendash 9$, $2015$ (South seed cohort into South garden), respectively. We chose times when we observed natural seedlings were of similar size. The North garden was located near natural populations along the Middle Fork of the Feather River ($39$° $46$\' $59.7$\" N, $120$° $38$\' $31.1$\" W, $1314$ mas, Plumas County, California, USA); the South garden was located along King Creek  ($32$° $54$\' $23.4$\" N, $116$° $37$\' $26.0$\" W, $1011$ mas, San Diego County, California, USA). These gardens are located close to two of the source populations (Fig. \ref{fig:range-map}). The North garden is $7.1$ km from the Little Jamison Creek source population; the South garden is $3.8$ km from Sweetwater River source population. Since \textit{M. cardinalis} is a riparian specialist, gardens were located within $50$ m of natural waterways where the microclimate closely resembled that of nearby natural populations. We irrigated plots with microperforated drip tape (Toro, Bloomington, Minnesota) placed about each seedling to help establish transplants and mimic riparian soil where \textit{M. cardinalis} naturally germinates. Seedlings were initially irrigated once per day. The duration of watering was adjusted at each garden to give each plant approximately the same volume of water. Once plants were established, we gradually reduced irrigation as evaporation declined while preventing drought stress. Irrigation ended when winter precipitation began. We discarded data on plants that died within the first month after transplanting, as this probably indicates transplant shock rather than natural mortality. We monitored reemergence in the spring using monthly censuses beginning on March $14$, $2016$ (South) and April $15$, $2016$ (North). If a plant was recorded in the last census of $2015$ and emerged in spring $2016$, we counted it as alive. If a plant did not reemerge, we recorded winter mortality.

@@

We estimated genotypic and population-level selection on winter survival as a function of germination rate. Long-term demographic studies of natural *M. cardinalis* populations use size- rather age-based models, so we do not know with certainty how important a fitness component winter survival is. However, population growth rate is moderately sensitive to the survival of prereproductive individuals of uncertain age; early survival is less important than recruitment, but more important than fecundity [@angert_demography_2006]. If genetic variation in germination rate within and between populations *causes* the probability of survival to change, then there should be significant genotypic selection on germination rate. Alternatively, populations with different mean germination rates may have different probabilities of survival as a consequence of other, confounding trait differences. If this is true, we predict a population-level correlation between germination rate and survival, but no evidence of genotypic selection on variation within populations. For each sample of breeding values and population averages from the posterior distribution, we used ordinary linear regression to calculate the relationship between germination rate and winter survival in both gardens at both genotypic and populational levels. We also calculated the variance among populations ($V_\text{pop}$) in germination rate and survival from the posterior distribution. We used the median and quantile intervals to estimate the slopes and quantify uncertainty in our estimates. We used the same approach to estimate quadratic selection coefficients as well.

@@

The vast majority of seeds germinated in $1 \textendash 2$ weeks, but northern source populations took a few days longer on average. $`r et["north", "Y"]`$ out of $`r et["north", "Y"] + et["north", "N"]`$ plugs ($`r germination_north`$%) had germinants in the North cohort; $`r et["south", "Y"]`$ out of $`r et["south", "Y"] + et["south", "N"]`$ plugs ($`r germination_south`$%) had germinants in the South cohort. South cohort seedlings also germinated about one day faster (mean = $`r round(as.numeric(days_summary[["days_to_emerge_mean"]][days_summary[["Cohort"]] == "south"]), 1)`$, median = $`r as.integer(days_summary[["days_to_emerge_median"]][days_summary[["Cohort"]] == "south"])`$) than the North cohort (mean = $`r round(as.numeric(days_summary[["days_to_emerge_mean"]][days_summary[["Cohort"]] == "north"]), 1)`$, median = $`r as.integer(days_summary[["days_to_emerge_median"]][days_summary[["Cohort"]] == "north"])`$). The estimated log-mean difference and 95% CIs in time to germinate in the North cohort was `r germ_north_diff`. After accounting for cohort and block effects, southern source populations consistently germinate faster (Fig. \ref{fig:mean-traits}a). For example, seeds from the slowest population (Rock Creek) germinated $`r sprintf("%.1f", ex[[1, 'd52']])`~(95\%~\text{CI}~[`r sprintf("%.1f", ex[[1, '.lower']])`,`r sprintf("%.1f", ex[[1, '.upper']])`])$ days slower than the fastest population (West Fork Mojave River). The discretized log-normal distribution was an adequate model of germination rate based on the similarity of the posterior predictions to the observed data (Fig. \ref{fig:pp_check}).

@@

Fewer plants survived over winter in the North garden than in the South garden (`r survival_north`% versus `r survival_south`%). In both gardens, southern source populations survived more often than northern source populations even after accounting for block effects (Fig. \ref{fig:mean-traits}b). For example, in the North garden, we estimated that plants originating from the most southern source population, Sweetwater River, survived `r p_CUR_north` of the time compared to `r p_RCK_north` for the most northern source population Rock Creek. Similarly, in the South garden the local population from Sweetwater River survived `r p_CUR_south` of the time compared to `r p_RCK_south` for Rock Creek. For most populations, the average temperature was warmer than the historical normal from $1961 \textendash 1990$, especially for the winter season in the South garden (Fig. \ref{fig:climate}). However, temperatures for the North garden were close to their historical norm.

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{figures/mean-traits.pdf}
  \caption{(Caption next page.)}
  \label{fig:mean-traits}
\end{figure}
\addtocounter{figure}{-1}
\begin{figure} [t!]
  \caption{(Previous page.) Germination rate and survival differ among source populations of \textit{Mimulus cardinalis}. Source populations are arrayed along the $x$-axis by latitude of origin from south to north. In both panels, the violins represent the posterior distribution of the population trait value and the bars are the median. Each black point is the median trait value for one of the individuals in the base population. Connecting letters above violins indicate populations that significantly different or not. The 95\% confidence interval for difference between populations includes 0 for those with the same letter. a. The northernmost populations germinate days later than southern populations in the same greenhouse environment. b. In both South (solid linetype around violin) and North (dashed linetype around violin) gardens, source populations originating from the south had higher probability of winter survival than northern source populations. However, the overall survival was greater in the South garden, such that even the northernmost source population (Rock Creek) had higher survival in the South than in the North garden.}
\end{figure}

## Variance in germination rate and survival within and between populations

@@

The genetic variation in germination rate within populations of *M. cardinalis* is similar to the genetic differences between them (Fig. \ref{fig:h2-germ}a). The difference between $V_\text{pop}$ and $V_G$ for the log-mean germination rate is `r diff_vpop_vg_germ`, but the confidence intervals include $0$. The environmental variance $V_E$ is greater than $V_G$, resulting in a moderate heritability $H^2$ (Fig. \ref{fig:h2-germ}b). Maternal and Block effects contributed little variance (Fig. \ref{fig:h2-germ}a). In contrast, there is virtually no genetic variance in winter survival ($p_\text{surv}$) in either garden, resulting in heritabilities that are effectively $0$ (Fig. \ref{fig:h2-surv}). There is some variation between populations, but it is substantially less than the unexplained environmental variance in both gardens (Fig. \ref{fig:h2-surv}). Parameter estimates and confidence intervals for all variance components and heritabilities are given in Table \ref{tab:vc_table_germ} for germination rate and Table \ref{tab:vc_table_surv} for winter survival.

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{figures/h2-germ.pdf}
  \caption{Estimated variance components and heritability of germination rate in units of log(days)$^2$. a. The variance in germination rate between source populations $V_\text{pop}$ is comparable to the genetic variation within populations $V_G$. The variance among Blocks and maternal parent ($V_M$) in the greenhouse is substantially lower whereas the unexplained environmental variance is higher $V_E$. b. This results a moderate heritability $H^2$ which is $V_G / V_P$. The point estimates are the median of the posterior distribution; thick lines are $80$\% confidence intervals; and thin lines are the $95$\% confidence intervals.}
  \label{fig:h2-germ}
\end{figure}

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{figures/h2-surv.pdf}
  \caption{Estimated variance components and heritability of winter survival in units of $p^2_\text{surv}$. The top row of facets are estimates from the North garden; the bottom row of facets are estimates from the South garden. On the $x$-axis, the source populations are arranged from left (orange) to right (blue) by latitude going from south to north. The right-most facet ($V_\text{pop}$ is grey because it is the variance among populations. On the $y$-axis (log$_{10}$-transformed for visual clarity) is the variance or heritability ($H^2$) depending on the facet. In both gardens, the unexplained environmental ($V_E$) is higher than the variance contributed by field Block, genetic variance ($V_E$), or variance between source populations ($V_\text{pop}$). Hence, the $H^2$ is very low in both gardens. The point estimates are the median of the posterior distribution; thick lines are $80$\% confidence intervals; and thin lines are the $95$\% confidence intervals.}
  \label{fig:h2-surv}
\end{figure}

## Directional selection favors faster germination between but not within populations

Since there was almost no genetic variance in winter survival in either garden, but some variance between populations, the only evidence for selection is between populations. Source populations originating from farther south in the species' range germinated faster and had higher winter survival than those originating for farther north (Fig. \ref{fig:selection}). However, within populations, genotypes that germinated faster did not have a different probability of survival. Selection coefficients and confidence intervals are given in Table \ref{tab:df_pop_lm} for between-population selection and Table \ref{tab:df_ind_lm} for genotypic selection. There was no evidence for stabilizing selection using quadratic regression (results not shown).

>>>>

---
title: 'Selection on early survival does not explain germination rate clines in *Mimulus cardinalis*'
author: "Christopher D. Muir$^{1,2,*}$, Courtney L. Van Den Elzen$^{1,3}$, Amy L. Angert$^{1,4}$"
output:
  bookdown::pdf_document2:
    includes:
      before_body: ../template/doc_prefix.tex
      in_header: ../template/preamble.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc: no
  bookdown::html_document2:
    number_sections: no
    theme: readable
    toc: yes
  bookdown::tufte_html2:
    number_sections: no
    toc: yes
  bookdown::word_document2: null
fontsize: 12pt
linestretch: 1.5
link-citations: yes
csl: ../template/american-journal-of-botany.csl
bibliography: ../ms/mimulus-germination.bib
---

@@

<!--- broad problem: local adaptation --->
Populations within species differ in ecologically important traits that often result from adaptation to different local environments [@turesson_genotypic_1922; @clausen_experimental_1948]. However, it is still rare that we understand the traits and selective agents underlying local adaptation [@wadgymar_identifying_2017]. The classic signature of local adaptation is crossing reaction norms for fitness measured in a reciprocal transplant experiment [@kawecki_conceptual_2004; @johnson_reciprocal_2021], such that populations have  higher relative fitness in their local environment but lower relative fitness in a foreign environment. For example, annual inland ecotypes of the yellow Monkeyflower *Mimulus guttatus* (Phrymaceae) have higher relative fitness in seasonally dry sites compared to perennial coastal ecotypes that are adapted to year-round water availability. The relative fitness of coastal and inland ecotypes is reversed on the coast because of life history and other genetic differences between ecotypes [@lowry_widespread_2010]. The prevalence of local adaptation implies that selection varies spatially, populations have heritable variation in fitness, and selection is stronger than migration or drift. Exceptional cases of local adaptation over short temporal or spatial scales [e.g. @wright_indirect_2013; @grant_40_2014; @richardson_microgeographic_2014; @lescak_evolution_2015; @barrett_linking_2019; @divittorio_natural_2020] likely contribute to an availability bias that leads many to perceive that local adaptation is ubiquitous. However, systematic meta-analyses reveal that local adaptation is often weak or nonexistent [@leimu_meta-analysis_2008; @hereford_quantitative_2009; @brady_understanding_2019] and may be getting weaker because of anthropogenic climate change [@bontrager_climate_2020]. Populations may not be locally adapted if differential selection is weak relative to migration or drift or they lack heritable variation in traits under selection. Alternatively, recent anthropogenic climate change may have erased the signature of local adaptation by causing some foreign populations to have higher fitness than local populations [@bontrager_climate_2020]. For example, in the alpine plant *Boechera strica*, simulated climate change (mimicking early snow melt) favors populations from lower elevations and snow addition restores the relative fitness advantage of high-elevation populations [@anderson_climate_2020]. Understanding what causes trait variation among populations within species, whether adaptive or not, will help biologists better predict how populations will respond to environmental change.

@@

There are several experimental methods for measuring the agents and strength of selection to different environments under natural conditions [@wadgymar_identifying_2017], but here we focus on differential selection. One key prediction is that the direction of selection and position of the phenotypic optimum will differ among environments. If genetic variation in a trait *causes* fitness to vary, then genotypic selection analyses [@rausher_measurement_1992] should find that genotypes with high trait expression are favored in one environment but disfavored in another. For example, selection favors higher seed dormancy in southern environments of *Arabidopsis thaliana* and lower dormancy in northern environments, presumably to optimally time seedling establishment with the onset of favorable summer conditions [@postma_early_2016]. Populations can also simultaneously differ as a result of both adaptive and non-adaptive processes. When locally adapted populations differ in traits that do not themselves confer local adaptation, we expect population-level covariation between divergent traits and fitness but do not expect within-population variation to correlate with fitness. Hence, comparing the direction of among-genotype and among-population selection can help distinguish between phenotypic clines that are important for local adaptation and those that may have evolved nonadaptively.

@@

*Mimulus cardinalis* [$=$*Erythranthe cardinalis* @lowry_case_2019] is a perennial forb native to the Western US (California and Oregon). It is self-compatible, predominantly outcrossing, and hummingbird pollinated. We used five focal populations from throughout the geographic range of *M. cardinalis* (Table \ref{tab:focal_pops}; Appendix S1, Fig. \ref{fig:range-map}; see the Supplementary Data with this article). These five focal populations are a subset of those we used previously to identify a cline in germination and other traits [@muir_grow_2017]. Seeds were collected in the field from mature, undehisced fruits left in open coin envelopes for $2 \textendash 4$ weeks to dry, then stored at room temperature. To minimize maternal effects, we grew a large number of field-collected seeds in the greenhouse and generated seed families for this experiment by hand-pollinating individuals using the breeding design described in the next section.

@@

$$V_G = 4~\sigma_{\mathrm{parent}}$$
Using $\sigma_{\mathrm{parent}}$ rather than $\sigma_{\mathrm{sire}}$ is more powerful because it uses all of the data. We estimated $\sigma_{\mathrm{parent}}$ and other parameters using a Bayesian mixed effects model. We describe the general approach here and provide detail specific to germination rate and survival below. We fit the model in *Stan* version $`r stringr::str_c(fit[["metadata"]]()["stan_version_major"], fit[["metadata"]]()["stan_version_minor"], fit[["metadata"]]()["stan_version_patch"], sep = ".")`$ [@stan_development_team_stan_2022] using **cmdstanr** version $`r packageVersion("cmdstanr")`$ [@gabry_cmdstanr_2022]. *Stan* calculates the posterior distribution of the model using the Hamiltonian Monte Carlo algorithm, which is similar to the more widely used Markov Chain Monte Carlo, but is faster and more efficient at sampling for many applications [@monnahan_faster_2017]. We used weakly informative normally distributed priors for parameters that affect the trait mean (intercepts and coefficients) and half-Cauchy priors for variance parameters. Weakly informative priors are strongly recommended for complex mixed models [@mcelreath_statistical_2016]. We ran $`r length(fit$metadata()$id)`$ parallel chains for $`r fit[["metadata"]]()[["iter_warmup"]]`$ warmup iterations, $`r fit[["metadata"]]()[["iter_sampling"]]`$ sampling iterations, and a thinning interval of $`r fit[["metadata"]]()[["thin"]]`$. This configuration allowed parameters to converge, which we defined as the convergence diagnostic $\hat{R} < 1.05$ [@vehtari_rank-normalization_2021]. We inspected posterior predictive plots to assess model adequacy using the `pp_check` function from the *R* package **bayesplot** version $`r packageVersion("bayesplot")`$ [@gabry_visualization_2019]. We used the posterior median for point estimates and calculated uncertainty with the $95$\% quantile intervals from the posterior distribution.

@@

We estimated germination rate under benign greenhouse conditions by measuring time to germination, the number of days to visible emergence from the soil. There were $48$ randomized blocks split evenly between two time cohorts (North cohort: August $3 \textendash 4$ and South cohort: August $22 \textendash 24$, $2015$) in the University of British Columbia greenhouse (Vancouver, Canada). The North seed cohort went into the North transplant garden (gardens described below) and the South seed cohort went into the South transplant garden. Seed cohorts were staggered so that seedlings were transplanted at the same ontogenetic stage, as described below. We distinguish between cohorts and gardens because germination for each transplant garden was assayed in the greenhouse prior to transplanting. Emergence is a reliable proxy for germination since seeds were sown directly on top of the soil, resulting in little delay between germination and visible cotyledons. We sowed $3 \textendash 5$ seeds per plug on moist perlite-peat moss potting medium (Sunshine #1, Sungro Horticulture, Agawam, Massachusetts, USA) and recorded the first day on which a plug had at least one germinated seedling. We later thinned each plug to one seedling by selecting the central-most individual, so the first germinated seedling recorded is generally not the same individual transplanted into the field. Hence, individual correlations between germination and survival are inappropriate and we analyze relationships between breeding values or population averages. We censused germination daily for most of the experiment, but census dates were more spread out later in the experiment when only a few plants remained. We accounted for census interval in our statistical model (see below). Gentle misting kept soil moist for the duration of the experiment. Most plugs ($`r length(which(data[["Germinate"]] == "Y"))`$ of $`r nrow(data)`$, $`r round(100 * length(which(data[["Germinate"]] == "Y")) / nrow(data), 1)`$%) had at least one germinant by the end of the experiment (North cohort: August 27, 2015; South cohort: September 20, 2015); we treated plants that did not germinate as missing data. $`r nRemove`$ ($`r round(100 * nRemove / length(which(data[["Germinate"]] == "Y")), 2)`$%) plants emerged before day $5$. We removed data on these individuals because they were likely not *M. cardinalis*, but rather contamination from another species' seed in the potting medium. Species can be difficult to tell apart as the cotyledons first emerge, but in plugs where a single germinant emerged before day $5$ it was always a contaminant species. It therefore seems most likely that all germinants before day $5$ were contamination, but observations during thinning showed that contamination was rare and it unlikely to have any substantial impact on our estimate of germination timing.

@@

After germination, we transplanted $20$ out of $24$ blocks of seedlings each to two experimental gardens located at the northern and southern portions of the species' range. We refer to these as the "North" and "South" gardens, respectively. We used some plants from the remaining $4$ blocks to replace those that did not germinate or died during transport. We transplanted seedlings at similar ontogenetic stages but temporally staggered. We planted the North and South cohorts $5 \textendash 6$ weeks after germination on September $9 \textendash 17$, $2015$ (North seed cohort into North garden) and October $2 \textendash 9$, $2015$ (South seed cohort into South garden), respectively. We chose times when we observed natural seedlings were of similar size. The North garden was located near natural populations along the Middle Fork of the Feather River ($39$° $46$\' $59.7$\" N, $120$° $38$\' $31.1$\" W, $1314$ mas, Plumas County, California, USA); the South garden was located along King Creek  ($32$° $54$\' $23.4$\" N, $116$° $37$\' $26.0$\" W, $1011$ mas, San Diego County, California, USA). These gardens are located close to two of the source populations (Appendix S1, Fig. \ref{fig:range-map}). The North garden is $7.1$ km from the Little Jamison Creek source population; the South garden is $3.8$ km from Sweetwater River source population. Since \textit{M. cardinalis} is a riparian specialist, gardens were located within $50$ m of natural waterways where the microclimate closely resembled that of nearby natural populations. We irrigated plots with microperforated drip tape (Toro, Bloomington, Minnesota, USA) placed about each seedling to help establish transplants and mimic riparian soil where \textit{M. cardinalis} naturally germinates. Seedlings were initially irrigated once per day. The duration of watering was adjusted at each garden to give each plant approximately the same volume of water. Once plants were established, we gradually reduced irrigation as evaporation declined while preventing drought stress. Irrigation ended when winter precipitation began. We discarded data on plants that died within the first month after transplanting, as this probably indicates transplant shock rather than natural mortality. We monitored reemergence in the spring using monthly censuses beginning on March $14$, $2016$ (South) and April $15$, $2016$ (North). If a plant was recorded in the last census of $2015$ and emerged in spring $2016$, we counted it as alive. If a plant did not reemerge, we recorded winter mortality.

@@

We estimated genotypic and population-level selection on winter survival as a function of germination rate. Long-term demographic studies of natural *M. cardinalis* populations use size- rather age-based models, so we do not know with certainty how important a fitness component winter survival is. However, population growth rate is moderately sensitive to the survival of prereproductive individuals of uncertain age; early survival is less important than recruitment, but more important than fecundity [@angert_demography_2006]. If genetic variation in germination rate within and among populations *causes* the probability of survival to change, then there should be significant genotypic selection on germination rate. Alternatively, populations with different mean germination rates may have different probabilities of survival as a consequence of other, confounding trait differences. If this is true, we predict a population-level correlation between germination rate and survival, but no evidence of genotypic selection on variation within populations. For each sample of breeding values and population averages from the posterior distribution, we used ordinary linear regression to calculate the relationship between germination rate and winter survival in both gardens at both genotypic and populational levels. We also calculated the variance among populations ($V_\text{pop}$) in germination rate and survival from the posterior distribution. We used the median and quantile intervals to estimate the slopes and quantify uncertainty in our estimates. We used the same approach to estimate quadratic selection coefficients as well.

@@

The vast majority of seeds germinated in $1 \textendash 2$ weeks, but northern source populations took a few days longer on average. $`r et["north", "Y"]`$ out of $`r et["north", "Y"] + et["north", "N"]`$ plugs ($`r germination_north`$%) had germinants in the North cohort; $`r et["south", "Y"]`$ out of $`r et["south", "Y"] + et["south", "N"]`$ plugs ($`r germination_south`$%) had germinants in the South cohort. South cohort seedlings also germinated about one day faster (mean = $`r round(as.numeric(days_summary[["days_to_emerge_mean"]][days_summary[["Cohort"]] == "south"]), 1)`$, median = $`r as.integer(days_summary[["days_to_emerge_median"]][days_summary[["Cohort"]] == "south"])`$) than the North cohort (mean = $`r round(as.numeric(days_summary[["days_to_emerge_mean"]][days_summary[["Cohort"]] == "north"]), 1)`$, median = $`r as.integer(days_summary[["days_to_emerge_median"]][days_summary[["Cohort"]] == "north"])`$). The estimated log-mean difference and 95% CIs in time to germinate in the North cohort was `r germ_north_diff`. After accounting for cohort and block effects, southern source populations consistently germinate faster (Fig. \ref{fig:mean-traits}a). For example, seeds from the slowest population (Rock Creek) germinated $`r sprintf("%.1f", ex[[1, 'd52']])`~(95\%~\text{CI}~[`r sprintf("%.1f", ex[[1, '.lower']])`,`r sprintf("%.1f", ex[[1, '.upper']])`])$ days slower than the fastest population (West Fork Mojave River). The discretized log-normal distribution was an adequate model of germination rate based on the similarity of the posterior predictions to the observed data (Appendix S1, Fig. \ref{fig:pp_check}).

@@

Fewer plants survived over winter in the North garden than in the South garden (`r survival_north`% versus `r survival_south`%). In both gardens, southern source populations survived more often than northern source populations even after accounting for block effects (Fig. \ref{fig:mean-traits}b). For example, in the North garden, we estimated that plants originating from the most southern source population, Sweetwater River, survived `r p_CUR_north` of the time compared to `r p_RCK_north` for the most northern source population Rock Creek. Similarly, in the South garden the local population from Sweetwater River survived `r p_CUR_south` of the time compared to `r p_RCK_south` for Rock Creek. For most populations, the average temperature was warmer than the historical normal from $1961 \textendash 1990$, especially for the winter season in the South garden (Appendix S1, Fig. \ref{fig:climate}). However, temperatures for the North garden were close to their historical norm.

## Variance in germination rate and survival within and among populations

@@

The genetic variation in germination rate within populations of *M. cardinalis* is similar to the genetic differences among them (Fig. \ref{fig:h2-germ}a). The difference between $V_\text{pop}$ and $V_G$ for the log-mean germination rate is `r diff_vpop_vg_germ`, but the confidence intervals include $0$. The environmental variance $V_E$ is greater than $V_G$, resulting in a moderate heritability $H^2$ (Fig. \ref{fig:h2-germ}b). Maternal and Block effects contributed little variance (Fig. \ref{fig:h2-germ}a). In contrast, there is virtually no genetic variance in winter survival ($p_\text{surv}$) in either garden, resulting in heritabilities that are effectively $0$ (Fig. \ref{fig:h2-surv}). There is some variation among populations, but it is substantially less than the unexplained environmental variance in both gardens (Fig. \ref{fig:h2-surv}). Parameter estimates and confidence intervals for all variance components and heritabilities are given in Appendix S1, Table \ref{tab:vc_table_germ} for germination rate and Appendix S1, Table \ref{tab:vc_table_surv} for winter survival.

## Directional selection favors faster germination among but not within populations

Since there was almost no genetic variance in winter survival in either garden, but some variance among populations, the only evidence for selection is among populations. Source populations originating from farther south in the species' range germinated faster and had higher winter survival than those originating for farther north (Fig. \ref{fig:selection}). However, within populations, genotypes that germinated faster did not have a different probability of survival. Selection coefficients and confidence intervals are given in Appendix S1, Table \ref{tab:df_pop_lm} for among-population selection and Appendix S1, Table \ref{tab:df_ind_lm} for genotypic selection. There was no evidence for stabilizing selection using quadratic regression (results not shown).
